---
title: "What do we really know about BCG?"
subtitle: "(from a mathematicians perspective)"
author: "Sam Abbott (@seabbs)"
institute: "University of Bristol"
date: "2016/06/14"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

class: inverse

# Outline

1. Background

1. Introduction to the BCG

1. Vaccine action and effectiveness

1. Duration of protection

1. Non-specific effects

1. Other uses

1. Summary

1. Signposting

---
class: inverse, center, middle


# Background

---

# Tuberculosis

* TB is primarily a respiratory disease (pulmonary TB) caused by the bacterium Mycobacterium tuberculosis, although it can also affect other parts of the body (extra-pulmonary TB)

* TB spreads via airborne droplets.  

* After infection 5-10% of individuals develop symptomatic TB within 2 years of infection.

* The majority of individuals enter a latent stage in which they passively carry TB mycobacterium.

* The chances of developing active disease and the severity of outcomes vary across age groups, with children being more likely to have severe outcomes.

* Risk factors include: positive HIV status; low socioeconomic status; high density living; homelessness; incarceration; and drug use.

---

# Global Tuberculosis<sup>1</sup> 

```{r, map-tb-globally, echo = FALSE, message = FALSE, fig.height = 5, fig.width = 10, dev = 'svg'}
getTBinR::map_tb_burden()
```

.footnote[
[1] Made with `getTbinR` (https://github.com/seabbs/getTBinR)
]

???

* Over 10 million new cases of active TB each year. 
???

---

# Global Tuberculosis<sup>1</sup> 

```{r, plot-tb-globally, echo = FALSE, message = FALSE, fig.height = 4.5, fig.width = 8, dev = 'svg'}
getTBinR::plot_tb_burden_summary(conf = NULL) +
  ggplot2::theme(legend.position = "right")
```

.footnote[
[1] Made with `getTbinR` (https://github.com/seabbs/getTBinR)
]

???

 
???
---

# Global Tuberculosis<sup>1</sup> 

```{r, plot-tb-annual-change-globally, echo = FALSE, message = FALSE, fig.height = 4.5, fig.width = 8, dev = 'svg'}
getTBinR::plot_tb_burden_summary(conf = NULL, annual_change = TRUE, facet = "Area") +
  ggplot2::theme(legend.position = "none")
```

.footnote[
[1] Made with `getTbinR` (https://github.com/seabbs/getTBinR)
]

???

* Incidence rates are falling globally but the rate of this decreases varies regionally. 

???
---

# Tuberculosis in the UK<sup>1</sup> 

```{r plot-tb-uk, echo = FALSE, fig.height = 5, fig.width = 10, dev = 'svg'}
plot <- tbinenglanddataclean::plot_historic_tb_ew()

suppressWarnings(print(plot))
```

.footnote[
[1] Made with `tbinenglanddataclean` (https://github.com/seabbs/tbinenglanddataclean)
]
---

class: inverse, center, middle


# The Bacillus Calmette–Guérin (BCG) vaccine

---

# History of the BCG

- Bacillus Calmette–Guérin (BCG) vaccine was first given to humans in 1921.

- It is a live vaccine and was developed by weakening a strain of Mycobacterium bovis, which is commonly found in cows, over a period of 13 years.

- Initially public acceptance was slow, with low take up until after the Second World War.

- Since then most counntries have introduced it into their routine vaccination schedules and it remains one of the most commonly administered vaccines.

- Howover, controversy remains and several countries have scaled back, or redirected, their usage in recent years.

---

# Usage

## WHO recommendations

## In the United Kingdom

---

class: inverse, center, middle


# Vaccine action, effectiveness and duration of protection

---

# Protection against active TB

---

# Protection from initial infection

---

# Variable effectiveness

---

# Duration of protection

---

class: inverse, center, middle

# Other uses

---

# Leprosy

---

# BCG for bladder cancer

---

# Non-specific effects of BCG

---


---

# Bovine TB

---

# Summary


---

class: inverse, center, middle

# Thanks for listening

---
